High-Dimensional Regularized Discriminant Analysis

### 2013-12-26 reports/paper-hdrda.Rnw
Lots of polish applied to the introduction. Much cleaner and more concise. Closes #10 and #15

The popular regularized discriminant analysis (RDA) classifier proposed by
\citeauthor{Friedman:1989tm} utilizes a biased covariance-matrix estimator that
partially pools the sample covariance matrices from linear and quadratic
disriminant analysis before shrinking the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure from a
grid of candidate values. We reformulate the RDA classifier in the
high-dimensional setting and then show that the classification decision rule is
equivalent to one in a subspace having a much lower dimension. As a result, the
dimension reduction involved yields a substantial reduction in computation
during model selection. Our parameterization also offers interpretability
previously lacking with the RDA classifier. We demonstrate with three artificial
data sets and four high-dimensional real data sets that our proposed classifier
is often superior to several recently proposed sparse and regularized
classifiers in terms of classification accuracy. Finally, we provide an
implementation of our proposed classifier in the {\tt sparsediscrim} R package,
available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers} to
construct automated pattern-recognition systems in a variety of disciplines,
such as medical research, facial recognition, remote sensing, and chemical
analysis \citep{TODO}. Despite the advances in classification to facilitate
technological advancements, well-established classifiers often degrade in
accuracy with modern, high-dimensional data sets when the number of features $p$
exceeds the total sample size $N$, due to the \emph{curse of dimensionality}
\citep{Bellman:1961tn}. For instance, linear discriminant analysis (LDA) and
quadratic discriminant analysis (QDA) are popular because of their simplicity,
speed of computation, and interpretability. However, both methods are
incalculable when $p > N$ because the class sample covariance matrices are
singular \citep*{Bouveyron:2007gx}.

As a result, researchers have emphasized three standard approaches to modifying
the \emph{LDA} and \emph{QDA} classifiers for application to high-dimensional
data: 1) covariance-matrix regularization, 2) dimension reduction, and 3) model
parsimony. Covariance-matrix regularization methods often are based on the
well-known ridge-regression approach of
\cite{Hoerl:1970cd}. \cite{Ching:2012fu}, \cite{Zhang:2010va}, \cite{Ye:2009gd},
\cite{Ji:2008wp}, \cite*{Guo:2007te}, \cite*{Srivastava:2007ud},
\cite{Ye:2006vx}, and \cite{Ye:2005uu} have proposed regularization methods to
improve the estimation of the class covariance matrices by adjusting the
eigenvalues of the sample covariance matrices to ensure positive
definiteness. Alternatively, dimension reduction typically decomposes a model to
orthogonal components and retains the components corresponding to the model's
intrinsic dimension. While some classificatory information may be lost,
dimension reduction methods have been shown to be effective in practice
\citep{YePaper,TODO}. Finally, \cite{Mai:2012bf}, \cite{Fan:2012iq},
\cite{Merchante:2012vk}, \cite{Clemmensen:2011kr}, \cite{Huang:2010ju}, and
\cite{Dudoit:2002ev} have proposed restrictions on the population covariance
matrices to reduce the number of parameters that must be estimated. For
instance, \cite{Dudoit:2002ev} have assumed the features within each class are
independent so that the population covariance matrices are diagonal, resulting
in positive-definite sample covariance matrices. For excellent overviews of the
regularization problem in discriminant analysis and other techniques, see
\cite{Murphy:2012uq} and \cite{Mkhadri:1997gy}.

We consider regularized discriminant analysis (\emph{RDA}) from a classic paper
by \cite{Friedman:1989tm}, who proposed improvements to the \emph{QDA}
classifier by employing a covariance matrix estimator that partially pools the
individual sample covariance matrices with the pooled sample covariance matrix
from the \emph{LDA} classifier. Given that the assumption of equal population
covariance matrices for the \emph{LDA} classifier is often too stringent in practice
\citep{Clemmensen:2011kr}, the \emph{RDA} classifier can relax the linearity
assumption by instead incorporating unequal covariance matrices while retaining
as special cases the \emph{LDA} and QDA classifiers. Covariance-matrix
regularization is then applied to the partially pooled covariance-matrix
estimator by shrinking it towards a scaled identity matrix to ensure positive
definiteness. We argue that the \emph{RDA} classifier inherently employs
dimension reduction in addition to regularization. After parameterizing the
\emph{RDA} classifier similar to \cite{Hastie:2008dt} and \cite{Halbe:2007ea},
we demonstrate that the intrinsic dimension of the \emph{RDA} classifier is
equal to the training sample size of each class.\footnote{TODO: Introduce HDRDA
  somewhere.} We then show that the \emph{RDA} decision rule is invariant to
adjustments of the zero eigenvalues. We show that our proposed classifier is
invariant to any adjustments of the eigenvalues corresponding to the null space
of the pooled sample covariance matrix. Applying a reasoning similar to
\cite{Ye:2006tq}, we decompose the decision rule of the \emph{HDRDA} classifier
and show that the sum related to the null space of the pooled sample covariance
matrix can be discarded. Hence, no gain in classification performance can be
obtained by adjusting these zero eigenvalues. This invariance property provides
a substantial increase in computational efficiency of model training and
selection. Moreover, the decision rule in the original feature space is
equivalent to a decision rule in a much lower dimension (i.e., the intrinsic
dimension), such that the matrix inverses and determinants can be rapidly
computed. This invariance property as well as the dual optimization problem
result in substantial reduction in computation. Additionally, our
parameterization of the \emph{RDA} classifier offers interpretability that has
been previously lacking \citep{Bensmail:1996fw}. We explicitly show that the
pooling parameter in the \emph{RDA} covariance-matrix estimator controls the
contribution of each training observation to the estimation of each class
covariance matrix. Furthermore, our parameterization includes a large family of
shrinkage methods, such as those from \cite{Srivastava:2007ww},
\cite{Rao:1971ul}, and other methods studied by \cite{Ramey:2013ji} and
\cite{Xu:2009fl}.

Given the \emph{RDA} classifier has been shown to have excellent performance in
the high-dimensional setting \citep{Webb:2011vu} but is limited by its
computationally intense model selection procedure, our work enables the
\emph{RDA} classifier to be employed with high-dimensional data in
practice. This is reassuring because the \emph{RDA} classifier remains widely
popular in the literature. In fact, variants of the RDA classifier have been
applied to microarray data \citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te},
facial recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep{Wu:1996us,Aeberhard:1993fp}.

After demonstrating that the \emph{HDRDA} classifier can be computed efficiently
and establishing its improved interpretability, we compare the computational
runtime of the \emph{HDRDA} classifier with an implementation of the \emph{RDA}
classifier from the {\tt klaR} R package to illustrate that the \emph{HDRDA}
classifier greatly reduces computation when $p \gg N$. We then study the
classification performance of the \emph{HDRDA} classifier on several artificial
and real high-dimensional data sets. We compare the \emph{HDRDA} classifier with
several recent several recent classifiers from \cite{Dudoit:2002ev},
\cite{Guo:2007te}, \cite{Pang:2009ik}, \cite{Tong:2012hw}, and
\cite{Witten:2011kc}, each of which is designed for small-sample,
high-dimensional data. We demonstrate that the \emph{HDRDA} classifier often
attains superior classification accuracy. We also include in our classification
study a support vector machine (\emph{SVM}) and $k$-nearest neigbors ($kNN$) as
benchmarks.

The remainder of this document is organized as follows. In Section
\ref{sec:preliminaries} we presents the discrimination problem and the necessary
notation to describe our contributions. In Section \ref{sec:rda} we present our
\emph{HDRDA} classifier along with its interpretation, properties, and
computationally efficient model-selection procedure. Section \ref{sec:sims}
describes our simulation studies of artificial and real data sets and also
describes the experimental results. We conclude with a brief discussion in
Section \ref{sec:discussion}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) = (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $ \bm x_i$ $(i =
1, \ldots, N)$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is, from
\eqref{eq:pooled-cov} we have
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$. We
can interpret \eqref{eq:sig-lambda-alternative2} as a covariance-matrix
estimator that borrows from $\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov} to
better estimate $ \bm \Sigma_k$. From \eqref{eq:sig-lambda-alternative2},
$\lambda$ weights the contribution of each of the $N$ observations in estimating
$\bm \Sigma_k$. Hence, for $0 < \lambda \le 1$, we estimate $\bm \Sigma_k$ with
$N$ rather than $n_k$ observations.\footnote{TODO: Add more to the
  interpretation here.}

@@

p <- ggplot(rda_weights, aes(lambda, Weight, linetype = Class, color = Class))
p <- p + geom_line(size = 1)
p <- p + theme_bw() + xlab(expression(lambda))
p <- p + theme(axis.text = element_text(size = 10))
p <- p + theme(strip.text.x = element_text(size = 16))
p <- p + theme(legend.title = element_text(size = 16, face = "bold"))
p <- p + theme(legend.text = element_text(size = 15))
p <- p + theme(legend.key.size = unit(1, "cm"))
p
@ 
\caption{Weights $c_{ik}(\lambda)$ of \emph{RDA} classifier as a function of the
  pooling parameter $\lambda$ when $n_1 = 50$, $n_2 = 25$, $n_3 = 15$, and $n_4$
  = 10.}
\label{fig:hdrda-weights}
\end{figure}

@@

>>>>

\citeauthor{Friedman:1989tm} proposed the popular regularized discriminant
analysis (RDA) classifier that utilizes a biased covariance-matrix estimator
that partially pools the sample covariance matrices from linear and quadratic
disriminant analysis before shrinking the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure from a
grid of candidate values. We reformulate the RDA classifier in the
high-dimensional setting and then show that the classification decision rule is
equivalent to one in a subspace having a much lower dimension. As a result, the
dimension reduction involved yields a substantial reduction in computation
during model selection. Our parameterization also offers interpretability
previously lacking with the RDA classifier. We demonstrate with three artificial
and four real high-dimensional data sets that our proposed classifier is often
superior to several recently proposed sparse and regularized classifiers in
terms of classification accuracy. Finally, we provide an implementation of our
proposed classifier in the {\tt sparsediscrim} R package, available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers} to
construct automated pattern-recognition systems in a variety of disciplines,
such as medical research, facial recognition, remote sensing, and chemical
analysis \citep{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

As a result, researchers have emphasized three standard approaches to modifying
the \emph{LDA} and \emph{QDA} classifiers for application to high-dimensional
data: 1) covariance-matrix regularization, 2) dimension reduction, and 3) model
parsimony. Covariance-matrix regularization methods are often based on the
well-known ridge-regression approach of
\cite{Hoerl:1970cd}. \cite{Ching:2012fu}, \cite{Zhang:2010va}, \cite{Ye:2009gd},
\cite{Ji:2008wp}, \cite*{Guo:2007te}, \cite*{Srivastava:2007ud},
\cite{Ye:2006vx}, and \cite{Ye:2005uu} have proposed regularization methods to
improve the estimation of the class covariance matrices by adjusting the
eigenvalues of the sample covariance matrices to ensure positive
definiteness. Alternatively, dimension reduction typically transforms the
original data in the high-dimensional space to a subspace having a much smaller
dimension such that redundant and noisy information is removed
\citep{Fan:2012iq,Ye:2009gd,Bouveyron:2007gx,Ye:2006vx,Antoniadis:2003bk}. Finally,
\cite{Mai:2012bf}, \cite{Fan:2012iq}, \cite{Merchante:2012vk},
\cite{Clemmensen:2011kr}, \cite{Huang:2010ju}, and \cite{Dudoit:2002ev} have
proposed restrictions on the population covariance matrices to reduce the number
of parameters that must be estimated. For instance, \cite{Dudoit:2002ev} have
assumed the features within each class are independent so that the population
covariance matrices are diagonal, resulting in positive-definite sample
covariance matrices. For excellent overviews of the regularization problem in
discriminant analysis and other techniques, see \cite{Murphy:2012uq} and
\cite{Mkhadri:1997gy}.

We consider regularized discriminant analysis (\emph{RDA}) from a classic paper
by \cite{Friedman:1989tm}, who proposed improvements to the \emph{QDA}
classifier by employing a covariance matrix estimator that partially pools the
individual sample covariance matrices with the pooled sample covariance matrix
from the \emph{LDA} classifier. Given that the assumption of equal population
covariance matrices for the \emph{LDA} classifier is often too stringent in
practice \citep{Clemmensen:2011kr}, the \emph{RDA} classifier's partial pooling
can relax the linearity assumption by instead incorporating unequal covariance
matrices while retaining as special cases the \emph{LDA} and QDA
classifiers. Covariance-matrix regularization is then applied to the partially
pooled covariance-matrix estimator by shrinking it towards a scaled identity
matrix to ensure positive definiteness.

We reparameterize the \emph{RDA} classifier similar to \cite{Hastie:2008dt} and
\cite{Halbe:2007ea} and refer to our parameterization as the high-dimensional
\emph{RDA} (\emph{HDRDA}) classifier. First, we show the pooling parameter in
the \emph{HDRDA} classifier determines the contribution of each training
observation to the estimation of each class covariance matrix, enabling
interpretability that has been previously lacking with the \emph{RDA} classifier
\citep{Bensmail:1996fw}. Furthermore, the \emph{HDRDA} classifier includes a
large family of shrinkage methods, such as those from \cite{Srivastava:2007ww},
\cite{Rao:1971ul}, and other methods studied by \cite{Ramey:2013ji} and
\cite{Xu:2009fl}. Next, applying a reasoning similar to \cite{Ye:2006tq}, we
show that the matrix operations corresponding to the null space of the pooled
sample covariance matrix are redundant and can be discarded from the
\emph{HDRDA} decision rule without loss of classificatory information. Moreover,
because the \emph{HDRDA} decision rule is invariant to adjustments to the zero
eigenvalues, the decision rule in the original feature space is equivalent to a
decision rule in a much lower dimension (i.e., the intrinsic dimension), such
that the matrix inverses and determinants of relatively small matrices can be
rapidly computed. Thus, the dimension reduction inherent in the \emph{HDRDA}
classifier yields a substantial reduction in computation during model selection.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier and then compare its computational runtime with an
implementation of the \emph{RDA} classifier from the {\tt klaR} R package to
illustrate that the \emph{HDRDA} classifier greatly reduces computation when $p
\gg N$. We then study the classification performance of the \emph{HDRDA}
classifier on three artificial and four real high-dimensional data sets. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite{Tong:2012hw}, \cite{Witten:2011kc},
\cite{Pang:2009ik}, \cite{Guo:2007te}, and \cite{Dudoit:2002ev}. We also include
as benchmarks support vector machines (\emph{SVM}) and $k$-nearest neigbors
($k$-NN) in our classification study.\footnote{TODO: If we change any of the
  classifiers in the simulations, they should be updated here.}

The remainder of this document is organized as follows. In Section
\ref{sec:preliminaries} we present the discrimination problem and the necessary
notation to describe our contributions. In Section \ref{sec:rda} we present the
\emph{HDRDA} classifier along with its interpretation, properties, and
computationally efficient model-selection procedure. Section \ref{sec:sims}
describes our simulation studies of artificial and real data sets and also
describes the experimental results. We conclude with a brief discussion in
Section \ref{sec:discussion}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) = (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $ \bm x_i$ $(i =
1, \ldots, N)$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is, from
\eqref{eq:pooled-cov} we have
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$. We
can interpret \eqref{eq:sig-lambda-alternative2} as a covariance-matrix
estimator that borrows from $\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov} to
better estimate $ \bm \Sigma_k$. From \eqref{eq:sig-lambda-alternative2},
$\lambda$ weights the contribution of each of the $N$ observations in estimating
$\bm \Sigma_k$. Hence, for $0 < \lambda \le 1$, we estimate $\bm \Sigma_k$ with
$N$ rather than $n_k$ observations.\footnote{TODO: Add more to the
  interpretation here and reference Figure \ref{fig:hdrda-weights}.}

@@

p <- ggplot(rda_weights, aes(lambda, Weight, linetype = Class, color = Class))
p <- p + geom_line(size = 1)
p <- p + theme_bw() + xlab(expression(lambda))
p <- p + theme(axis.text = element_text(size = 10))
p <- p + theme(strip.text.x = element_text(size = 16))
p <- p + theme(legend.title = element_text(size = 16, face = "bold"))
p <- p + theme(legend.text = element_text(size = 15))
p <- p + theme(legend.key.size = unit(1, "cm"))
p
@ 
\caption{Weights $c_{ik}(\lambda)$ of \emph{RDA} classifier as a function of the
  pooling parameter $\lambda$ when $n_1 = 50$, $n_2 = 25$, $n_3 = 15$, and $n_4$
  = 10. TODO: Scale down figure.}
\label{fig:hdrda-weights}
\end{figure}

@@

Given the \emph{RDA} classifier has been shown to have excellent performance in
the high-dimensional setting \citep{Webb:2011vu} but is limited by its
computationally intense model selection procedure, our work enables the
\emph{RDA} classifier to be employed with small-sample, high-dimensional data in
practice. This is reassuring because the \emph{RDA} classifier remains widely
popular in the literature. In fact, variants of the RDA classifier have been
applied to microarray data \citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te},
facial recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep{Wu:1996us,Aeberhard:1993fp}.
